Keep your Laziness in Check

### 2018-03-09 paper/main.tex
Refactoring a bit in the rot example

To describe the laziness behavior, we need to first introduce a few
types that describe how lists are used. We use the
term \textit{demand} to refer to the shape of such usage. Concretely,
a \textit{demand} is a shape that describes the used part of a data
structure. For lists, \textit{demands} will be prefixes of lists. We
use the type \lk{ListD} to represent such demands.
%
\begin{inlinecode}
  data ListD a = NilD
               | ConsD (Thunk a) (Thunk (ListD a))
\end{inlinecode}
%
The \lk{Thunk} type used in \lk{ListD} has 2 constructors: \lk{E}
for \textit{evaluated} and \lk{T}, a nullary constructor,
for \textit{thunk}. In \lk{ListD}, \lk{Thunk} wraps around all the
list constructor fields, and this allows us to represent lack of
further evaluation by using the \lk{T} constructor. For example, given
the list \lk{[1, 2, 3]}, the value \lk{ConsD T (ConsD T T)} represents
the \textit{demand} that reveals the first 2 \lk{Cons} constructors of
the list, but does not evaluate any other parts of the list. To
improve readability of demand values, we will also use the
syntax \lk{\_ : \_ : \_} instead of explicit demand type values. In
this expression, the first two underscores represent unevaluated
elements in the list, while the last underscore represents unevaluated
tail of the list.

@@

Putting these pieces together, we can specify the laziness behavior
of \lk{rotate} with Okasaki's insight: as long as the front list is
longer than the back list, demanding the first element from the result
of \lk{rotate} only causes evaluation of the first element of the
front list ~\cite{Okasaki95simpleand}. Since \lk{rotate} is always
used by supplying the empty list \lk{[]} as the accumulator argument,
we define \lk{rot fs bs = rotate fs bs []}, and show a specification
of \lk{rot}.
%
\begin{inlinecode}
rot_spec :: Shaped a => Spec '[[a], [a]] [a]
rot_spec =
  Spec $ \predict d fs bs ->
    let (fs', bs') = splitAt (length fs) (cap d)
        spineLen  = length (cap (d ++ [undefined]))  -- # of spine thunks forced
        overflow  = spineLen       > length fs  -- begun taking from bs?
        overrot   = length (cap d) > length bs  -- forced all of bs?
        padLength =
          length bs `min`
            if overflow
            then length bs - length bs'
            else length (cap d)
        spinePad = replicate padLength thunk
    in predict
         (                    fs' ++ if overflow            then [] else thunk)
         (spinePad ++ reverse bs' ++ if overflow || overrot then [] else thunk)
\end{inlinecode}
%
The \lk{rotate\_spec} exactly captures the core idea from Okasaki's \lk{rotate}
operation, and testing the reverse and append implementation against this
specification immediately produces a counter example.
\begin{figure}[H]
\centering
\includegraphics[width=0.8\textwidth]{revappend_counterexample}
\caption{Counterexample for \lk{revAppend} under \lk{rot\_spec}}
\end{figure}

>>>>

To describe the laziness behavior, we need to first introduce a few of
StrictCheck's types. First of all, we use a type \lk{Thunk} to capture
whether a particular field is forced or not.
%
\begin{inlinecode}
  data Thunk = E !a | T
\end{inlinecode}
%
Similarly to a \lk{Maybe}, the \lk{Thunk} datatype has two
constructors: a nullary one \lk{T} corresponding to thunks and a
strict single-argument one \lk{E} corresponding to evaluated values.

Secondly, StrictCheck introduces the notion of a \textit{demand}: a
shape that describes the used part of a data structure. For
lists, \textit{demands} will be prefixes of lists. We use the
type \lk{ListD} to represent such list demands.
%
\begin{inlinecode}
  data ListD a = NilD
               | ConsD (Thunk a) (Thunk (ListD a))
\end{inlinecode}
%
In \lk{ListD}, \lk{Thunk} wraps around all the list constructor
fields, and this allows us to represent lack of further evaluation by
using the \lk{T} constructor.

\leo{The following is slightly too dense. How can we take out some
part of this as code?}
For example, given the list \lk{[1, 2, 3]}, the value \lk{ConsD T
(ConsD T T)} represents the \textit{demand} that reveals the first
2 \lk{Cons} constructors of the list, but does not evaluate any other
parts of the list. To improve readability of demand values, we will
also use the syntax \lk{\_ : \_ : \_} instead of explicit demand type
values. In this expression, the first two underscores represent
unevaluated elements in the list, while the last underscore represents
unevaluated tail of the list.
%
We also use metavariables \lk{\X{i}} to denote either a thunk or an
evaluated field. For instance, \lk{\X1 : \X2 : \_} can denote the
above 2-\lk{Cons} demand \lk{\_ : \_ : \_}, a similar demand where
both cons cells are evaluated \lk{ 1 : 2 : \_}, or a mix of both.

@@

Putting these pieces together, we can specify the laziness behavior
of \lk{rotate}. Since \lk{rotate} is always used by supplying the
empty list \lk{[]} as the accumulator argument, we define \lk{rot fs
bs = rotate fs bs []}, and show a specification of \lk{rot}.  Before
giving the actual specification however, let's try to build up some
intuition a series of example demands for the simpler, reverse and
append version:
%
\begin{inlinecode}
  rot_simple fs bs = fs ++ reverse bs
\end{inlinecode}

All of our example demands will be of the following form.
%
\rotex{[1,2,3]}{[6,5,4]}{[1,2,3,4,5,6]}
      {\X1 : \X2 : \_}{\X1 : \X2 : \_}{\_}
%
The fist column shows the input lists \lk{fs} and \lk{bs}; here they
are \lk{[1,2,3]} and \lk{[4,5,6]} respectively. The second column
shows the result of \lk{rot\_simple fs bs} when fully evaluated;
here \lk{[1,2,3,4,5,6]}. The third column shows the actual output
demand exerted; just like we described above, the demand \lk{\X1 : \X2
: \_} means that we only force the first two cons cells and, depending
on whether the \lk{\X{i}} are thunks or not, we might also force the
first two elements. The first and third column together are the actual
inputs to our specification. The second column is there for the
readers convenience.

The last column is the demand on the inputs predicted by our
specification.  In this first example, where the output demand is
smaller than the input list, the demand on the first argument \lk{fs}
is equal to the demand, while the demand on the second list \lk{bs} is
the empty demand. Indeed, if concretely the output demand is \lk{ \_ :
2 : \_} (which could for example be obtained by printing \lk{hd \$
tl \$ rot\_simple [1,2,3] [6,5,4]}), then only the first
two \lk{Cons}-cells of \lk{fs} will be forced, and of those only the
second one will have its integer field evaluated.

On the other hand, if the output demand is larger than the first list,
then \textit{all} of the spine of second argument has to be evaluated.
%
\rotex{[1,2,3]}{[6,5,4]}{[1,2,3,4,5,6]}
      {\X1 : \X2 : \X3 : \X4 : \_}{ [ \X1 , \X2 , \X3 ] } { [ \_ , \_ , \X4 ] }
%
The predicted demand on the first list will be the top part of the
demand of length equal to \lk{length fs}. The predicted demand on the
second list will be the reverse of the remaining demand, padded with
thunks if necessary.
\leo{One more tricky case, when the demand on the output forces
the null constructor of the bs list. How do we talk about that?}

Now we can give the actual StrictCheck specification
of \lk{rot\_simple}:

\leo{Debug the stupid spec and add it and describe it}

We can now move on to the more complex \lk{rot} function from Okasaki:
as long as the front list is longer than the back list, demanding the
first element from the result of \lk{rotate} only causes evaluation of
the first element of the front list ~\cite{Okasaki95simpleand}.
%
\begin{inlinecode}
rot_spec :: Shaped a => Spec '[[a], [a]] [a]
rot_spec =
  Spec $ \predict d fs bs ->
    let (fs', bs') = splitAt (length fs) (cap d)
        spineLen  = length (cap (d ++ [undefined]))  -- # of spine thunks forced
        overflow  = spineLen       > length fs  -- begun taking from bs?
        overrot   = length (cap d) > length bs  -- forced all of bs?
        padLength =
          length bs `min`
            if overflow
            then length bs - length bs'
            else length (cap d)
        spinePad = replicate padLength thunk
    in predict
         (                    fs' ++ if overflow            then [] else thunk)
         (spinePad ++ reverse bs' ++ if overflow || overrot then [] else thunk)
\end{inlinecode}
%
The \lk{rotate\_spec} exactly captures the core idea from Okasaki's \lk{rotate}
operation, and testing the reverse and append implementation against this
specification immediately produces a counter example.
\begin{figure}[H]
\centering
\includegraphics[width=0.8\textwidth]{revappend_counterexample}
\caption{Counterexample for \lk{revAppend} under \lk{rot\_spec}}
\end{figure}
